Comparing the Direction and Degree of Selection on Mutations Between SARS-CoV-2 Phylogenetic Clades

### 2022-11-03 notebooks/expected_and_actual_counts_ratio.Rmd
Modify and simplify code to only handle two clades

# Create a list of the clades to include in comparisons.
clades <- c("20A","20B","20C")

@@

```{r filter_all_data}
# Include data from certain countries AND exclude branches with very high mutation rates and questionable mutations.
filtered_SARS_mutation <- subset(SARS_mutation, subset==country & exclude=="FALSE" & expected_count>=min_expected)

# Add column with the mutation type (clade founder.mutation).
filtered_SARS_mutation$mutation_type <- paste(str_sub(filtered_SARS_mutation$nt_mutation,1,1), ".", str_sub(filtered_SARS_mutation$nt_mutation,-1,-1), sep="")
```

@@

@@

```{r stacked_bar, echo=FALSE}
# Stacked bar graph for the mutation counts normalized by the expected mutation rate for each clade
ggplot(filtered_SARS_mutation, aes(fill=mutation_type, y=actual_count/expected_count, x=clade))+
  geom_bar(position="stack", stat="identity")+
  theme_bw()+
  labs(x="Phylogenetic Clade", y="Normalized Mutation Count")
```

@@

```{r nucleotide_site, echo=FALSE}
ggplot(filtered_SARS_mutation, aes(color=mutation_type, y=log(actual_count/expected_count), x=nt_site))+
  geom_point()+
  theme_bw()+
  labs(x="", y="")
```

```{r INSERT, echo=FALSE}
#ggplot(filtered_SARS_mutation, aes(fill=INSERT, y=INSERT, x=INSERT))+
  #geom_tile()+
  #theme_bw()+
  #labs(x="", y="")
```

@@

```{r subset_by_clade}
# Create a subset of the data using only the clades that are going to be compared.
clade_SARS_mutation <- filtered_SARS_mutation[filtered_SARS_mutation$clade %in% clades,]
```

Combine nucleotide site, mutation type and clade to create a reference column in the data set. This column will be referenced to obtain observed and expected counts for clade comparisons. Additionally, remove columns from the data set not necessary for contingency table analyses or creating volcano plots.

```{r filter_contingency_data}
# Filter mutation data for Fisher's Exact test, and combine nucleotide sites, mutation types and clades to make an iterative reference.
fisher_data <- data.frame(nt_mutation_clade=paste(clade_SARS_mutation$nt_site, clade_SARS_mutation$mutation_type, clade_SARS_mutation$clade, sep="_"), observed=clade_SARS_mutation$actual_count, expected=clade_SARS_mutation$expected_count, observed_expected=clade_SARS_mutation$observed_expected)
```

Create a two-column matrix of all possible clade comparisons, each column listing one clade in the comparison. This matrix should have [n x (n-1)]/2 rows where n is the number of clades.

```{r clade_comparisons}
# Create a two-column matrix with all of the possible clade-clade comparisons (each row contains one comparison).
clade_comparisons <- combinations(n=length(clades), r=2, v=clades, repeats.allowed=FALSE)
```

For each mutation type (e.g., A -> T) at each nucleotide site, this notebook will generate contingency table analyses for each comparison in the set of clade comparisons. To expand the clade comparison matrix to include mutation types at nucleotide sites, generate a unique list of each mutation type (i.e., three mutations for one clade founder) at each nucleotide site.

```{r unique_mutations}
# Combine nucleotide sites and mutation types, and filter list to only include unique nucleotide mutations.
nt_mutation_list <- unique(paste(clade_SARS_mutation$nt_site, clade_SARS_mutation$mutation_type, sep="_"))
```

Expand both the clade comparison matrix and the mutation list to create a combined matrix of all possible clade comparisons for each mutation type at each nucleotide site. To do this, repeat the clade comparison matrix by the number of unique mutations, and repeat each mutation in the mutation list by the number of clade comparisons.

```{r combined_clade_comparisons}
# In preparation for combining clade comparisons with unique nucleotide mutations, repeat the nucleotide mutation list by the number of clade comparisons.
rep_nt_mutation <- data.frame(nt_mutation=rep(nt_mutation_list, each=nrow(clade_comparisons)))

# In preparation for combining clade comparisons with unique nucleotide mutations, repeat the clade comparison matrix by the number of unique nucleotide mutations.
rep_clades <- do.call("rbind", replicate(length(nt_mutation_list), clade_comparisons, simplify=FALSE))

# Combine the nucleotide mutations to each column in the clade comparison matrix to create a reference for the fisher data.
comparisons <- data.frame(clade1=paste(rep_nt_mutation$nt_mutation, rep_clades[,1], sep="_"), clade2=paste(rep_nt_mutation$nt_mutation, rep_clades[,2], sep="_"))
```

Not every mutation will exist in each clade. Therefore, remove comparisons where observed and expected count data does not exist for one or both clades in the comparison.

```{r filter_for_comparisons}
# Drop comparisons where the nucleotide site, mutation type, clade reference does not exist in the observed-expected data frame.
# DOUBLE CHECK THIS: NOT CONVINCED IT IS WORKING
filtered_comparisons <- comparisons[comparisons$clade1 %in% fisher_data$nt_mutation_clade & comparisons$clade2 %in% fisher_data$nt_mutation_clade,]
```

To decrease the amount of time it takes to run the contingency tests, add columns to the filtered comparisons data frame with the observed counts, expected counts and observed/expected ratios for clade 1 and clade 2.

```{r contingency_matrix}
# Create an ID to maintain the order of rows (and thus comparisons) after merging data frames.
filtered_comparisons$id <- c(1:nrow(filtered_comparisons))

# Merge each of the clades in the comparison matrix with the data to align observed, expected and observed/expected values.
clade1 <- merge(filtered_comparisons[,c("id","clade1")], fisher_data[,c("nt_mutation_clade","observed","expected","observed_expected")], by.x="clade1", by.y="nt_mutation_clade")
clade2 <- merge(filtered_comparisons[,c("id","clade2")], fisher_data[,c("nt_mutation_clade","observed","expected","observed_expected")], by.x="clade2", by.y="nt_mutation_clade")

# Add observed, expected and observed/expected data to the comparison matrix.
{filtered_comparisons$clade1_observed <- clade1[order(clade1$id),"observed"]
filtered_comparisons$clade1_expected <- clade1[order(clade1$id),"expected"]
filtered_comparisons$clade2_observed <- clade2[order(clade2$id),"observed"]
filtered_comparisons$clade2_expected <- clade2[order(clade2$id),"expected"]
filtered_comparisons$clade1_obs_exp <- clade1[order(clade1$id),"observed_expected"]
filtered_comparisons$clade2_obs_exp <- clade2[order(clade2$id),"observed_expected"]}
```

@@

```{r p_value_correction}
# Identify clades from the two columns in the comparison matrix and combine into one column.
comparison_p_values$clades <- data.frame(clades=paste(str_split(comparison_p_values$clade1, "_", simplify=TRUE)[,3], str_split(comparison_p_values$clade2, "_", simplify=TRUE)[,3], sep="."))

# Count the number of rows for each clade (i.e., number of hypothesis tests).
clade.table <- table(comparison_p_values$clades)

# Add a column for the number of hypothesis tests conducted for each pair of clades.
comparison_p_values$num_tests <- apply(comparison_p_values$clades, 1, FUN=function(c){clade.table[[c[1]]]})

# Using Bonferroni Correction, adjust the p-values by the number of hypothesis tests.
comparison_p_values$p_values_corrected <- comparison_p_values$p_value/comparison_p_values$num_tests
```

@@

```{r fold-change}
# Calculate fold change for plotting volcano plots [clade 1 observed/clade 1 expected]/[clade 2 observed/clade 2 expected].
comparison_p_values$fold_change <- foldchange(comparison_p_values$clade1_obs_exp, comparison_p_values$clade2_obs_exp)
```

@@

```{r volcano_plot}
# Only plot data for the two clades being compared.
plot_data <- comparison_p_values[which(comparison_p_values$clades=="20B.20C"),]
ggplot(plot_data, aes(x=fold_change, y=-log10(p_value)))+
  geom_point()
```

@@

```{r verify_analyses}
# sanity check analyses (distribution of synonymous p-values)

```

>>>>

# Identify the two clades that will be included in comparisons.
clades <- c("20A","22B")

@@

```{r filter_all_data}
# Include data from certain countries AND exclude branches with very high mutation rates and questionable mutations.
filtered_SARS_mutation <- subset(SARS_mutation, subset==country & exclude=="FALSE" & expected_count>=min_expected)
```

@@

The Fisher's Exact test can only accept integers, so round the expected counts prior to running analyses (i.e., gives control over how values are rounded).

```{r round_expected}
# Round expected counts because contingency table analyses (i.e., Fisher's Exact test) can only analyze integers
filtered_SARS_mutation$rounded_expected <- round(filtered_SARS_mutation$expected_count)
```

@@

```{r stacked_bar, echo=FALSE}
# Add column with the mutation type (clade founder.mutation).
add_mutation <- filtered_SARS_mutation
add_mutation$mutation_type <- paste(str_sub(add_mutation$nt_mutation,1,1), ".", str_sub(add_mutation$nt_mutation,-1,-1), sep="")

# Stacked bar graph for the mutation counts normalized by the expected mutation rate for each clade
ggplot(add_mutation, aes(fill=mutation_type, y=actual_count/expected_count, x=clade))+
  geom_bar(position="stack", stat="identity")+
  theme_bw()+
  labs(x="Phylogenetic Clade", y="Normalized Mutation Count")
```

@@

```{r nucleotide_site, echo=FALSE}
ggplot(add_mutation, aes(color=mutation_type, y=log(actual_count/expected_count), x=nt_site))+
  geom_point()+
  theme_bw()+
  labs(x="", y="")
```

@@

```{r subset_by_clade}
# Create a subset of the data using only the clades that are going to be compared.
filtered_clades <- filtered_SARS_mutation[filtered_SARS_mutation$clade %in% clades,]
```

```{r restructure_data}
# Create two data frames with data for clade 1 and clade 2.
clade1 <- filtered_clades[which(filtered_clades$clade==clades[1]),c("nt_mutation","clade","actual_count","rounded_expected","observed_expected","synonymous","four_fold_degenerate")]
clade2 <- filtered_clades[which(filtered_clades$clade==clades[2]),c("nt_mutation","clade","actual_count","rounded_expected","observed_expected","synonymous","four_fold_degenerate")]

# Change column names so that clade 1 and clade 2 data frames can be combined
colnames(clade1)[2:ncol(clade1)] <- paste("clade1", colnames(clade1)[2:ncol(clade1)], sep="_")
colnames(clade2)[2:ncol(clade2)] <- paste("clade2", colnames(clade2)[2:ncol(clade2)], sep="_")
colnames(clade1)[which(colnames(clade1)=="clade1_clade")] <- "clade1"
colnames(clade2)[which(colnames(clade2)=="clade2_clade")] <- "clade2"

# Merge the two data frames by mutation so that each row represents one comparison.
filtered_comparisons <- merge(clade1, clade2, by="nt_mutation")
```

@@

```{r p_value_correction}
# Using Bonferroni Correction, adjust the p-values by the number of hypothesis tests.
comparison_p_values$p_value_corrected <- comparison_p_values$p_value/nrow(comparison_p_values)
```

@@

```{r fold-change}
# Calculate fold change for plotting volcano plots [clade 1 observed/clade 1 expected]/[clade 2 observed/clade 2 expected].
# Double check calculation with Jesse's paper
comparison_p_values$fold_change <- log2(comparison_p_values$clade1_observed_expected/comparison_p_values$clade2_observed_expected)
```

@@

```{r volcano_plot}
# Only plot data for the two clades being compared.
ggplot(comparison_p_values, aes(x=fold_change, y=-log10(p_value_corrected)))+
  geom_point()+
  theme_bw()
```

@@

```{r verify_analyses}
# sanity check analyses (distribution of synonymous p-values)
verify_analyses <- comparison_p_values
verify_analyses <- comparison_p_values[which(comparison_p_values$clade1_four_fold_degenerate=="TRUE"),]
nb <- sqrt(nrow(verify_analyses))
bw <- (max(verify_analyses$p_value_corrected)-min(verify_analyses$p_value_corrected))/nb
ggplot(verify_analyses, aes(x=p_value_corrected))+
  geom_histogram(bins = nb, binwidth = bw)
```

```{r alternative, echo=FALSE}
# Script that can handle 2+ clades: 1.380048, 1.151285, 1.034135 minutes
# Script that can only handle 2 clades: 50.01184, 58.11915, 58.6249 seconds

# Filter mutation data for Fisher's Exact test, and combine nucleotide sites, mutation types and clades to make an iterative reference.
#filtered_clades$nt_mutation_clade <- paste(filtered_clades$nt_mutation, filtered_clades$clade, sep="_")

# Create a two-column matrix with all of the possible clade-clade comparisons (each row contains one comparison).
#clade_comparisons <- combinations(n=length(clades), r=2, v=clades, repeats.allowed=FALSE)

# In preparation for combining clade comparisons with unique nucleotide mutations, repeat the nucleotide mutation list by the number of clade comparisons.
#rep_nt_mutation <- data.frame(nt_mutation=rep(unique(filtered_clades$nt_mutation), each=nrow(clade_comparisons)))

# In preparation for combining clade comparisons with unique nucleotide mutations, repeat the clade comparison matrix by the number of unique nucleotide mutations.
#rep_clades <- do.call("rbind", replicate(length(unique(filtered_clades$nt_mutation)), clade_comparisons, simplify=FALSE))

# Combine the nucleotide mutations to each column in the clade comparison matrix to create a reference for the fisher data.
#comparisons <- data.frame(clade1=paste(rep_nt_mutation$nt_mutation, rep_clades[,1], sep="_"), clade2=paste(rep_nt_mutation$nt_mutation, rep_clades[,2], sep="_"))

# Drop comparisons where the nucleotide site, mutation type, clade reference does not exist in the observed-expected data frame.
#filtered_comparisons <- comparisons[(comparisons$clade1 %in% filtered_clades$nt_mutation_clade & comparisons$clade2 %in% filtered_clades$nt_mutation_clade),]

# Create an ID to maintain the order of rows (and thus comparisons) after merging data frames.
#filtered_comparisons$id <- c(1:nrow(filtered_comparisons))

# Merge each of the clades in the comparison matrix with the data to align observed, expected and observed/expected values.
#clade1 <- merge(filtered_comparisons[,c("id","clade1")], filtered_clades[,c("nt_mutation_clade","actual_count","rounded_expected","expected_count","observed_expected","synonymous","four_fold_degenerate")], by.x="clade1", by.y="nt_mutation_clade")
#clade2 <- merge(filtered_comparisons[,c("id","clade2")], filtered_clades[,c("nt_mutation_clade","actual_count","rounded_expected","expected_count","observed_expected","synonymous","four_fold_degenerate")], by.x="clade2", by.y="nt_mutation_clade")

# Change column names so that clade 1 and clade 2 data frames can be combined
#colnames(clade1)[3:ncol(clade1)] <- paste("clade1", colnames(clade1)[3:ncol(clade1)], sep="_")
#colnames(clade2)[3:ncol(clade2)] <- paste("clade2", colnames(clade2)[3:ncol(clade2)], sep="_")

# Order clade 1 and clade 2 data frames by id and combine
#fisher_data <- cbind(clade1[order(clade1$id),], clade2[order(clade2$id),])
#fisher_data[,which(colnames(fisher_data) == "id")] <- NULL

# Identify clades from the two columns in the comparison matrix and combine into one column.
#comparison_p_values$clades <- data.frame(clades=paste(str_split(comparison_p_values$clade1, "_", simplify=TRUE)[,2], str_split(comparison_p_values$clade2, "_", simplify=TRUE)[,2], sep="."))

# Count the number of rows for each clade (i.e., number of hypothesis tests).
#clade.table <- table(comparison_p_values$clades)

# Add a column for the number of hypothesis tests conducted for each pair of clades.
#comparison_p_values$num_tests <- apply(comparison_p_values$clades, 1, FUN=function(c){clade.table[[c[1]]]})
```
